---
title: "MB5370 Module 04. Workshop 1 - Introduction"
author: "Nicole Welch"
output: html_document
date: "2024-03-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Github
``` {r git}
#Creating a github token
usethis::create_github_token()

# Introducing myself to Git
library(usethis)
use_git_config(user.name = "nicolewelch22", user.email = "welchn22@gmail.com")

#install.packages("usethis")
credentials::git_credential_ask()
usethis::git_sitrep()

# Entering token as credential 
gitcreds::gitcreds_set()

#  Getting Git and GitHub talking to each other (Pushing)
usethis::use_github()

```

# Install and load tidyverse packages
```{r tidyverse}
## install.packages("tidyverse") # Delete this line once installed
library("tidyverse")
```


# Load the data (mpg is built into ggplot2)
```{r loading data}
data(mpg)
```


# Quick data checks
```{r data check}
head(mpg)
glimpse(mpg)
summary(mpg)
```


# Create first plot
```{r plot}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = disp, y = hwy))


#The plot shows a negative relationship between engine size (displ) and fuel efficiency (hwy). In other words, cars with big engines use more fuel. What does this say about fuel efficiency and engine size?
```

# Understanding grammar of graphics
```{r grammar}
ggplot() 
# Or with just the data specified?
ggplot(data = mpg)
```

# When you’re creating a plot, you essentially need two attributes of a plot: a geom and aesthetics.
```{r aesthetics}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

# Change point color by class:
```{r color by class}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

# Change point size by class:
```{r size by class}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
# Warning: Using size for a discrete variable is not advised. 
```

# Change transparency (alpha) by class
```{r alpha}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

# Change point shape by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```


# Make all points pale turqouise
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "paleturquoise")
```


# Troubleshooting 
```{r troubleshoot}
# ggplot(data = mpg) 
# + geom_point(mapping = aes(x = displ, y = hwy)) 
# the + should be on top line
```

# Faceting
```{r facets}
# For discrete variables, facet_wrap(), ~ dictates which variable you want to subset your data with 
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)

# facet with more then one variable, facet_grid(), Two variables using ~ to split them 
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)

# use a . if you do not want to facet in the rows or column dimension
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl)
```

# Lines
```{r smooth lines}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

# straight line
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))

# changing geom type in my plot
ggplot(data = mpg) + 
  #  geom_point(mapping = aes(x = displ, y = hwy)) # points horrible 
  geom_smooth(mapping = aes(x = displ, y = hwy)) # try smooth line, 'fixed mappings'

# change line type 
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))

# setting group aesthetic to a categorical variable to draw multiple objects
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))

# Changing color based on drv value
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,
  )

# Notes: 
  # So let’s recap. A geom is an object that your plot uses to represent the data. To change the geom type in your plot, simply change the geom function that you add to your plot template. Sometimes you may want to try a few things out, in which case you could use comments to help you remember what worked and what didn’t. 

```  

# Multiple geoms
```{r}
#Showing underlying data and how it relates to the geom_smooth line
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))

# more efficient than code above, easier to change a variable that you are plotting 
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

# using mappings under ggplot to reduce duplication in code, helpful for changing point styles or the development of your own customization
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()

# Specifying different data for each layer 
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

# Transformations and stats

## Load data
```{r load data}
data("diamonds")
glimpse(diamonds)
```

# Bar plot
```{r bar plot}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

# Overriding defaults
```{r overriding defaults}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")

# What if we want to know the proportion rather than the absolute frequency?
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))

#Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.
#If so, can use # geom_bar(mapping = aes(x = cut, y = stage(after_stat = prop), group = 1))
```


# Plotting statistical details with error bars
```{r error bars}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```


# Aesthetic adjustments adjustments
```{r aesthetics}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut)) # border
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut)) # fill

# color by another variable
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))

# position adjustments to customize your plots with raw data, change heights, and dodge
# To alter transparency (alpha)

# To alter transparency (alpha)
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")

# To color the bar outlines with no fill color
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")

# Stacking, out of a count of 1
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")

# Overlapping object directly beside one another 
ggplot(data = diamonds) + 
geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")

# 'Jitter' adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatterplots but not barplots.
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```